---
title: "stat_test_part-2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pressure, echo=FALSE}
library(tidyverse)
library(ggpubr)
```

## Hypothesis Testing

## Considering diff_index

### Which test to use?

paired/unpaired?
- we are using two random variables: diff_index affiliated to Republicans and diff_index associated to Democrats.
- Republicans and Democrats are two different populations
-> unpaired

parametric/non parametric?
- data distrbution is not normal
-> non parametric test

data type?
- diff_index is metric variable
-> Wilcoxon Rank Sum test

### Hypothesis

for an unpaired test:

two options, null hypothesis of means (1) and null hypothesis of comparisons(2)

(1): no diff in mean between the two samples. 
(2): P(diff_index_dem<diff_index_dem) = P(diff_index_dem>diff_index_dem)

I think we should go for (1) as (2) seems more appropriated for ordinal variable


### Code

ONCE data_fin IN THE WORKING ENVIRONMENT


```{r pressure, echo=FALSE}
# splitting the dataset in two 
sub_dem <- data_fin[data_fin$'party' == 'D', ]
sub_rep <- data_fin[data_fin$'party' == 'R', ]
```

```{r pressure, echo=FALSE}
# diff_index distribution for Democrats
gghistogram(sub_dem$diff_index, bins = 6, fill= 'steelblue', xlab = 'Difficulty index', main = 'Difficulty Index: Democrats')
```


```{r pressure, echo=FALSE}
# diff_index distribution for Republicans
gghistogram(sub_rep$diff_index, bins = 6, fill= 'steelblue', xlab = 'Difficulty index',  main = 'Difficulty Index: Republicans')

```

```{r pressure, echo=FALSE}

# is R ranking automatically the values?
#-> yes: same value when calculating:
#   wilcox.test(sort(sub_dem$diff_index), sort(sub_rep$diff_index), alternative = "two.sided", paired = FALSE, conf.int = TRUE)
#   wilcox.test(sub_dem$diff_index, sub_rep$diff_index, alternative = "two.sided", paired = FALSE, conf.int = TRUE)

# alt="two.sided"?
# conf.level = ... ? what is the confidence interval we would like?
# do not need and exact p value -> normal approximation of the p value (by default) is used then
# conf.int = TRUE to get : "median of the difference between a sample x and a sample from y"
# effect size r = Z/ sqrt(N), to be calculated manually


# what is the wilcox.test argument that differentiate a wilcoxon test under mean hypothesis than wilcoxon test under comparison hypothesis?

wilcox.test(sub_dem$diff_index, sub_rep$diff_index, alternative = "two.sided", paired = FALSE, conf.int = TRUE)
# not warning saying that the sample is to small to get a high confidence interval -> good
```

## Considering diff_subj

### Which test to use?

paired/unpaired?
- we are using two random variables: diff_subj affiliated to Republicans and diff_index associated to Democrats.
- Republicans and Democrats are two different populations
-> unpaired

parametric/non parametric?
- data distrbution is not normal
-> non parametric test

distributions:
-distri of diff_index_rep and diff_index_dem are really similar 

data type?
- diff_subj is binary /not metric
->  Wilcoxon again
->  independent sample t test?

### Hypothesis

null hypothesis of comparisons
H0:P(diff_subj_dem<diff_subj_dem) = P(diff_subj_dem>diff_subj_dem)

