---
title: 'UCB W203 - Lab 1: Hypothesis Testing'
author: "Kevin Lustig, Rebecca Nissan, Anuradha Passan, Giorgio Soggiu"
date: "3/1/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages install}

knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(stringr)
library(tidyverse)
library(scales)
library(readxl)
library(rstatix)
```

# Part 1: Foundational Exercises

## 1.1 Professional Magic

### 1.1.1 Type I Error of the test - Rebecca
The type I error rate (i.e. false positive) is the probability of rejecting the null hypothesis whe it is correct. The type I error would be the probability of getting 0 or 6, with the assumption that p = 0.5 (null). 

This will be the alpha 
```{r}

```

### 1.1.2 Power of test given p = 0.75 - Rebecca 

```{r}

```

## 1.2 Wrong Test, Right Data - Kevin
In the Likert scale, the meaningful distance between the different scale points is not consistent. That is, assuming the Likert scale for the websites survey includes five points from 1 = "Very Unsatisfied" to 5 = "Very Satisfied," with 3 being "Neutral," we cannot say that a change from 1 to 3 and from 2 to 4 are equivalent quantifiable changes in opinion ^[At least, not with only five scale points; see, e.g.: Huiping Wu and Shing-On Leung, “Can Likert Scales Be Treated as Interval Scales?—a Simulation Study,” Journal of Social Service Research 43, no. 4 (June 2017): pp. 527-532, https://doi.org/10.1080/01488376.2017.1329775.]. However, in fact the change in quality of experience necessary for a given respondent to go from Very Unsatisfied with one site to Neutral with the other may in fact be considerably less than the change needed to go from Unsatisfied to Satisfied, though these each consist of a difference of two steps. Therefore, though the values produced are numeric, these data violate one of the assumptions for a paired t-test -- the use of metric, rather than ordinal, data. 

A paired t-test relies on metric data because, like other related tests including the z-test, it is fundamentally a calculation of the difference of means between reference groups. A paired t-test would ask of our survey data: is the mean difference between paired opinion scores different than what we would expect if there were no preference for either website (mean difference within pairs = 0)? Stated otherwise, on average across all respondents, how likely is it that there is really a preference for one site or the other, and how large a preference? However, because of the aforementioned limitation of Likert scale values, we cannot meaningfully parse a mean paired disparity of e.g., +2, because to calculate this requires assuming that non-comparable changes from any one Likert scale point to another are equivalent. The mean of the paired differences is thus meaningless. It is even difficult to trust the directionality of the mean difference across all pairs (respondents like the mobile website more or less than the regular website without regard to how much), as in calculating a mean value purely from the raw Likert scale scores, we may calculate an incorrect value by not correctly taking into account the "weights" of the differences of opinion in, again, Very Unsatsified and Neutral versus Unsatisfied and Satisfied. It's possible to conceive of a scenario in which even the sign of the mean difference is therefore incorrect. 

It is this last point on directionality that suggests an alternative approach to this issue. A non-parametric paired sign test allows us to analyze our ordinal data provided the observations are independent and identically distributed. It does not attempt, like the t-test, to quantify the size of the difference in opinion within pairs, if any. Rather, it treats all positive changes in opinion as equivalent, and likewise with all negative changes. This alternative test has two main drawbacks. First, it does not have the statistical power of a paired t-test. Second, it loses substantial information present in the original survey responses in the form of the exact values within each paired set of responses. However, in doing so, it allows us to avoid the inaccurate mean calculation of the t-test, and focus on a more accurate analysis of a simpler question: do respondents prefer one website over the other? In looking solely at increases or decreases in opinion score, the paired sign test therefore gives us a reasonable expectation of finding such an effect if one exists in the data. 

## 1.3 Test Assumptions - Giorggio and Annie 

### 1.3.1 World Happiness 
Scenario: We have two variables: Life.Ladder and Log.GDP.per.Capita, and we want to see whether countries in high GDP per capita are more or less happy than people in countries with low GDP per capita. 

Assumptions for two-sample t-test 
1. Normal - CHECK! 
2. i.i.d. - yes (assuming same data collection and data transformation procuedures), explain 
3. Metric variables - yes, explain 

https://www.datanovia.com/en/lessons/normality-test-in-r/#check-normality-in-r 

```{r}
# Load in dataset 
wh_data_raw <- read.csv('datasets/happiness_WHR.csv')

# Select necessary variables 
wh_data <- wh_data_raw %>% select(Life.Ladder, Log.GDP.per.capita)

# Get a quick look at the two relevant variables 
summary (wh_data$Life.Ladder)
summary(wh_data$Log.GDP.per.capita)

# Calculate the mean GDP per capita (will be useful in sorting high and low GDP per capita countries)
gdp_capita_mean <- mean(wh_data$Life.Ladder, na.rm = TRUE)
gdp_capita_mean

# Split the data into low and high gdp per capita countries 
## But first remove rows that have NAs 
wh_data_fin <- na.omit(wh_data)
## Now split into the high and low gdp per capita country groups
high_gdp_cap <- wh_data_fin %>% filter(Log.GDP.per.capita > gdp_capita_mean)
high_gdp_cap

low_gdp_cap <- wh_data_fin %>% filter(Log.GDP.per.capita < gdp_capita_mean)
low_gdp_cap

# Observe the histograms of both variables 
hist(high_gdp_cap$Life.Ladder) #make pretty 

#hist(low_gdp_cap$Life.Ladder) #fix #make pretty 

# Can do a Shapiro test as an extra source of evidence to test for normalcy
shapiro_test1 <- wh_data_fin %>% shapiro_test(Life.Ladder, Log.GDP.per.capita)
shapiro_test1_results <- as.data.frame(shapiro_test1)
```

### 1.3.2 Legislators 

Scenario: We want to test whether Democratic or Republic senators are older, with two variables party and age (age needs to be calculated from DOB). 

Assumptions for Wilcoxon Rank-Sum test:
1. Ordinal variables - yes, explain 
2. i.i.d. - yes, explain 
3. Same shape and spread of the the two variables - box whisker plot, histogram 

```{r}
leg_data <- read.csv('datasets/legislators-current.csv')

# Need to calculate the age for each legislator 

# Check whether the variables have the similar shape/spread - box whisker plot or histogram 

# Check for some metric on similarities between the two variable distributions


summary(legislators_current$'birthday')
summary(legislators_current$'type')
#str(birth_date)  to get the format of the dates

sub_dem <- legislators_current[legislators_current$'type' == 'sen', ]
sub_rep <- legislators_current[legislators_current$'type' == 'rep', ]

today_date <- as.Date("2022-02-20")

birth_date_dem <- as.numeric(difftime(today_date, as.Date(sub_dem$'birthday'), unit = "weeks"))/52.25 
birth_date_rep <- as.numeric(difftime(today_date, as.Date(sub_rep$'birthday'), unit = "weeks"))/52.25 

hist(birth_date_dem)
hist(birth_date_rep)
```

### 1.3.3 Wine and Health 

Scenario: We want to test whether these countries have more deaths from heart disease or liver disease.

Assumptions for Wilcoxon Signed Rank Test:
1. Metric Variables - yes, explain 
2. i.i.d. - yes, explain 
3. Difference is symmetric - need to test for this
4. Paired - explain why/why not
```{r}
# install.packages("wooldridge")
library(wooldridge)
wine_data  <- wine

# test for symmetry 

install.packages("wooldridge")
library(wooldridge)
# ?wine

summary(wine)
diff_hearth_liver <- as.numeric(wine$heart - wine$liver)
hist(diff_hearth_liver)

```

### 1.3.4 Attitudes Toward the Religion 

Scenario: We would like to know whether the U.S. population feels more positive towards Protestants or Catholics. 

Assumptions for a Paired t-Test
1. Metric Variables - yes, explain 
2. i.i.d. - yes, explain 
3. Paired - yes, explain 
4. Normalcy - Check for this 
```{r}
rel_data <- read.csv('datasets/GSS_religion.csv')

# Check for normalcy with histograms, and can do also the Shapiro test 

summary(GSS_religion)
hist(GSS_religion$prottemp)
hist(GSS_religion$cathtemp)
```

# Part 2: Statistical Analysis

Data source: 2020 American National Election Studies (ANES) - 2020 Time Series Study 

Research question: Did Democratic voters or Republican voters experience more difficulty voting in the 2020 election?

Define the following:
1. Democratic voters
2. Republican voters
3. Factors contributing to difficulty in voting --> create an index?

Data cleaning and wrangling - Kevin 

Form final dataset 

EDA and basic information on the variables and proof we can make assumptions needed for tests we will run

Run tests

Interpret tests

Create visualizations (if not done before)

Write report


```{r}
anes_data <- read.csv('datasets/anes_timeseries_2020_csv_20220210.csv')
```