---
title: "stat_test_part-2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pressure, echo=FALSE}
library(tidyverse)
library(rstatix)
library(ggpubr)
```

## Hypothesis Testing

## Considering diff_index

### Which test to use?

paired/unpaired?
- we are using two random variables: diff_index affiliated to Republicans and diff_index associated to Democrats.
- Republicans and Democrats are two different populations
-> unpaired

parametric/non parametric?
- data distrbution is not normal
-> non parametric test

data type?
- diff_index is categorical/ordinal variable
-> Wilcoxon Rank Sum test

### Hypothesis

H0: The diff index is the same for Rep and Dem 

### Code

ONCE data_fin IN THE ENVIRONMENT


```{r pressure, echo=FALSE}
# splitting the dataset in two 
sub_dem <- data_fin[data_fin$'party' == 'D', ]
sub_rep <- data_fin[data_fin$'party' == 'R', ]
```

```{r pressure, echo=FALSE}
# diff_index distribution for Democrats
gghistogram(sub_dem$diff_index, bins = 6, fill= 'steelblue', xlab = 'Difficulty index', main = 'Difficulty Index: Democrats')
```


```{r pressure, echo=FALSE}
# diff_index distribution for Republicans
gghistogram(sub_rep$diff_index, bins = 6, fill= 'steelblue', xlab = 'Difficulty index',  main = 'Difficulty Index: Republicans')

```

alt="two.sided"?
conf.level = ... ? what is the confidence interval we would like?

```{r pressure, echo=FALSE}
wilcox.test(sub_dem$diff_index, sub_rep$diff_index, alternative = "two.sided", paired = FALSE)
```

## Considering diff_subj

### Which test to use?

paired/unpaired?
- we are using two random variables: diff_index affiliated to Republicans and diff_index associated to Democrats.
- Republicans and Democrats are two different populations
-> unpaired

parametric/non parametric?
- data distrbution is not normal
-> non parametric test

data type?
- diff_subj is binary
->  Wilcoxon again
->  t test?
