---
title: 'UCB W203 - Lab 1: Hypothesis Testing'
author: "Kevin Lustig, Rebecca Nissan, Anuradha Passan, Giorgio Soggiu"
date: "3/1/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages install}

knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(stringr)
library(tidyverse)
library(scales)
library(readxl)
library(rstatix)
library(ggpubr)
```

# Part 1: Foundational Exercises

## 1.1 Professional Magic

### 1.1.1 Type I Error of the test - Rebecca
The type I error rate (i.e. false positive) is the probability of rejecting the null hypothesis whe it is correct. The type I error would be the probability of getting 0 or 6, with the assumption that p = 0.5 (null). 

This will be the alpha 
```{r}

```

### 1.1.2 Power of test given p = 0.75 - Rebecca 

```{r}

```

## 1.2 Wrong Test, Right Data - Kevin
In the likert scale the "distance" between the different options is not consistent, thus violating the metric scale assumption needed to run a paired t-test. We could suggest a paired sign test where the only only assumptions needed are: ordinal variables and i.i.d. 

## 1.3 Test Assumptions - Giorggio and Annie 

### 1.3.1 World Happiness 
Scenario: We have two variables: Life.Ladder and Log.GDP.per.Capita, and we want to see whether countries in high GDP per capita are more or less happy than people in countries with low GDP per capita. 

Assumptions for two-sample t-test 
1. Normal - CHECK! 
2. i.i.d. - yes (assuming same data collection and data transformation procuedures), explain 
3. Metric variables - yes, explain 

https://www.datanovia.com/en/lessons/normality-test-in-r/#check-normality-in-r 

```{r}
# Load in dataset 
wh_data <- read.csv('datasets/happiness_WHR.csv')

# Select necessary variables 
wh_data <- wh_data %>% select(Life.Ladder, Log.GDP.per.capita)

# Get a quick look at the two relevant variables 
summary (wh_data)

# Calculate the mean GDP per capita (will be useful in sorting high and low GDP per capita countries)
gdp_capita_mean <- mean(wh_data$Life.Ladder, na.rm = TRUE)

# Split the data into low and high gdp per capita countries 
## But first remove rows that have NAs 
wh_data <- na.omit(wh_data)
## Now split into the high and low gdp per capita country groups
high_gdp_cap <- wh_data %>% filter(Log.GDP.per.capita > gdp_capita_mean)


low_gdp_cap <- wh_data %>% filter(Log.GDP.per.capita < gdp_capita_mean)


# Observe the histograms of both variables 
hist(high_gdp_cap$Life.Ladder) #make pretty 

#hist(low_gdp_cap$Life.Ladder) #fix #make pretty 

# Can do a Shapiro test as an extra source of evidence to test for normalcy
shapiro_test1 <- wh_data %>% shapiro_test(Life.Ladder, Log.GDP.per.capita)
```

### 1.3.2 Legislators 

Scenario: We want to test whether Democratic or Republic senators are older, with two variables party and age (age needs to be calculated from DOB). 

Assumptions for Wilcoxon Rank-Sum test:
1. Ordinal variables - yes, explain 
2. i.i.d. - yes, explain 
3. Same shape and spread of the the two variables - box whisker plot, histogram 

```{r}
# Load in Data set 
leg_data <- read.csv('datasets/legislators-current.csv')

# Select necessary variables and calculate age 
leg_data  <- leg_data %>% 
  select(birthday, party, type) %>%
  filter(type == 'sen') %>%
  mutate(age = as.numeric(difftime(Sys.Date(), as.Date(birthday), unit = "weeks"))/52.25)
leg_data <- leg_data %>% select(party, age)

# Split the data by party
dem_sen <- leg_data %>% filter(party == 'Democrat')
rep_sen <- leg_data %>% filter(party == 'Republican')
indep <- leg_data %>% filter(party == 'Independent') 
## We see all Senators have been accounted for as the number of rows of above 3 data frames adds up to 100 

# Get a quick look at the relevant variables (Democrats and Republicans only) 
summary(dem_sen$age)
summary(rep_sen$age)

# Check whether the variables have the similar shape/spread - box whisker plot or histogram 
hist(dem_sen$age)
hist(rep_sen$age)
# Check for some metric on similarities between the two variable distributions
```

### 1.3.3 Wine and Health 

Scenario: We want to test whether these countries have more deaths from heart disease or liver disease.

Assumptions for Wilcoxon Signed Rank Test:
1. Metric Variables - yes, explain 
2. i.i.d. - yes, explain 
3. Difference is symmetric - need to test for this
4. Paired - explain why/why not
```{r}
# Load in dataset 
library(wooldridge)
wine_data  <- wine

# Select necessary variables
wine_data <- wine_data %>% 
  select(country, heart, liver)

# Test for symmetry
## First calculate the difference between Heart and Liver 
wine_data <- wine_data %>% mutate(difference = heart - liver)
## Look at the histogram too see if there is symmetry 
gghistogram(wine_data, x = 'difference', y = '..density..', fill= 'steelblue', bins = 5, add_density = TRUE)

```

### 1.3.4 Attitudes Toward the Religion 

Scenario: We would like to know whether the U.S. population feels more positive towards Protestants or Catholics. 

Assumptions for a Paired t-Test
1. Metric Variables - yes, explain 
2. i.i.d. - yes, explain 
3. Paired - yes, explain 
4. Normalcy - Check for this 
```{r}
# Load in dataset 
rel_data <- read.csv('datasets/GSS_religion.csv')

# Select necessary variables 
rel_data <- rel_data %>% select(cathtemp, prottemp)

# Get a quick look at the relevant variables 
summary(rel_data)

#Observe the histograms for both variables 
gghistogram(rel_data$cathtemp, fill= 'steelblue', bins = 5)
gghistogram(rel_data$prottemp, fill= 'steelblue', bins = 5)

# Conduct a Shapiro test to also help determine normalcy 
shapiro_test2 <- rel_data %>% shapiro_test(cathtemp, prottemp)
```

# Part 2: Statistical Analysis

Data source: 2020 American National Election Studies (ANES) - 2020 Time Series Study 

Research question: Did Democratic voters or Republican voters experience more difficulty voting in the 2020 election?

Define the following:
1. Democratic voters
2. Republican voters
3. Factors contributing to difficulty in voting --> create an index?

Data cleaning and wrangling - Kevin 

Form final dataset 

EDA and basic information on the variables and proof we can make assumptions needed for tests we will run

Run tests

Interpret tests

Create visualizations (if not done before)

Write report


```{r}
anes_data <- read.csv('datasets/anes_timeseries_2020_csv_20220210.csv')
```