---
title: 'UCB W203 - Lab 1: Hypothesis Testing'
author: "Kevin Lustig, Rebecca Nissan, Anuradha Passan, Giorgio Soggiu"
date: "3/1/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1: Foundational Exercises

## 1.1 Professional Magic

### 1.1.1 Type I Error of the test - Rebecca
The type I error rate (i.e. false positive) is the probability of rejecting the null hypothesis whe it is correct. The type I error would be the probability of getting 0 or 6, with the assumption that p = 0.5 (null). 

This will be the alpha 
```{r}

```

### 1.1.2 Power of test given p = 0.75 - Rebecca 

```{r}

```

## 1.2 Wrong Test, Right Data - Kevin
In the likert scale the "distance" between the different options is not consistent, thus violating the metric scale assumption needed to run a paired t-test. We could suggest a paired sign test where the only only assumptions needed are: ordinal variables and i.i.d. 

## 1.3 Test Assumptions - Giorgio and Annie 

### 1.3.1 World Happiness 
Scenario: We have two variables: Life.Ladder and Log.GDP.per.Capita, and we want to see whether countries in high GDP per capita are more or less happy than people in countries with low GDP per capita. 

Assumptions for two-sample t-test 
1. Normal - CHECK! 
2. i.i.d. - yes (assuming same data collection and data transformation procuedures), explain 
3. Metric variables - yes, explain 

```{r}
wh_data <- read.csv('datasets/happiness_WHR.csv')

# Calculate the mean GDP per capita 

# Split the data into low and high gdp per capita countries 

# Check out histograms of both variables 

# Can do a Shapiro test as an extra source of evidence to test for normalcy
# https://www.datanovia.com/en/lessons/normality-test-in-r/#check-normality-in-r
```

### 1.3.2 Legislators 

Scenario: We want to test whether Democratic or Republic senators are older, with two variables party and age (age needs to be calculated from DOB). 

Assumptions for Wilcoxon Rank-Sum test:
1. Ordinal variables - yes, explain 
2. i.i.d. - yes, explain 
3. Same shape and spread of the the two variables - box whisker plot, histogram 

```{r}
leg_data <- read.csv('datasets/legislators-current.csv')

# Need to calculate the age for each legislator 

# Check whether the variables have the similar shape/spread - box whisker plot or histogram 

# Check for some metric on similarities between the two variable distributions
```

### 1.3.3 Wine and Health 

Scenario: We want to test whether these countries have more deaths from heart disease or liver disease.

Assumptions for Wilcoxon Signed Rank Test:
1. Metric Variables - yes, explain 
2. i.i.d. - yes, explain 
3. Difference is symmetric - need to test for this
4. Paired - explain why/why not
```{r}
# install.packages("wooldridge")
library(wooldridge)
wine_data  <- wine

# test for symmetry 

```

### 1.3.4 Attitudes Toward the Religion 

Scenario: We would like to know whether the U.S. population feels more positive towards Protestants or Catholics. 

Assumptions for a Paired t-Test
1. Metric Variables - yes, explain 
2. i.i.d. - yes, explain 
3. Paired - yes, explain 
4. Normalcy - Check for this 
```{r}
rel_data <- read.csv('datasets/GSS_religion.csv')

# Check for normalcy with histograms, and can do also the Shapiro test 
```

# Part 2: Statistical Analysis

Data source: 2020 American National Election Studies (ANES) - 2020 Time Series Study 

Research question: Did Democratic voters or Republican voters experience more difficulty voting in the 2020 election?

Define the following:
1. Democratic voters
2. Republican voters
3. Factors contributing to difficulty in voting --> create an index?

Data cleaning and wrangling - Kevin 

Form final dataset 

EDA and basic information on the variables and proof we can make assumptions needed for tests we will run

Run tests

Interpret tests

Create visualizations (if not done before)

Write report


```{r}
anes_data <- read.csv('datasets/anes_timeseries_2020_csv_20220210.csv')
```